%----------------------------------------------------------------------------------------
%	PART - Using the Screen and Keyboard
%----------------------------------------------------------------------------------------

\makeatletter\@openrightfalse
\part{Using the Screen and Keyboard}

\chaptertypein{
	\keybackgroundcolor{gray}
	\keytextcolor{black}
	10 PRINT "\shiftkey\clrhomekey"\\
	20 SLEEP 10\\
	30 PRINT "your name"\\
	40 SLEEP 10\\
	50 GOTO 10
}

\newpage
\partheading{Using the Screen and Keyboard}

This chapter assumes that you've read and understand Chapter 1: {\emph Getting
to Know Your Commander X16}.  If you not, go back and read it.\\

Now that you have written your first program in BASIC on the Commander X16,
it's time to look at the simplest thing you can make the X16 do: displaying
characters\footnote{A 'character' is any symbol that can be displayed on the
screen, such as letters, numbers, punctuation, and the graphic symbols that can
be typed} to the screen.  As you've already seen, the {\ttfamily PRINT}
statement can be followed by a series of character surrounded by quotation
marks, causing the series of charcters to be displayed to the screen.  This
series of characters is called a {\em string}.  Let's write a program to print
a string to the screen.\\

It's nice to start a new program on an empty screen, so lets clear the screen
and place the cursor at the top by holding down the \shiftkey key and pressing
the \clrhomekey key.  To tell the Commander X16 that you are writing a new
program, type \keybackgroundcolor{white}\key{n}\key{e}\key{w} and then press
the \returnkey key.  Now you can begin to type the following program, paying
attention to line number and punctuation marks:\\

\key{1}\key{0} \spacebar \key{p}\key{r}\key{i}\key{n}\key{t} \spacebar
\key{"}\key{h}\key{e}\key{l}\key{l}\key{o}\key{,} \spacebar
\key{w}\key{o}\key{r}\key{l}\key{d}\key{!}\key{"} \returnkey

\note {

	The \key{,} key is just the \doublekey{<\\,} key pressed {\em without}
	holding the \shiftkey key, and the \key{!} key is just the \doublekey{!\\1}
	key pressed {\em while} holding the \shiftkey key.

}

Now that your program is typed into the X16, run it by typing
\key{r}\key{u}\key{n} and pressing \returnkey.

You should see the following printed to the next line:

\codeblock {
	HELLO, WORLD!\\
}

After your program finishes printing out the message, it should print a blank
line, a line with the {\ttfamily READY} prompt, and finally place your blinking
cursor on the following line.  Your screen should look something like this:\\

\begin{center}
	\screenbox{2.75in}{2in}{

		NEW\\
		
		READY.\\
		10 PRINT "HELLO, WORLD"\\
		RUN\\
		HELLO, WORLD!\\

		READY.\\
		\cursor

	}
\end{center}

When you typed {\ttfamily NEW} and hit \returnkey, the Commander X16 also left
a blank line followed by the {\ttfamily READY} prompt, just as it did when your
own program ran.  You can think of BASIC commands like {\ttfamily NEW} as being
built-in programs, and they act much in the same way as the programs you will
write.\\

%----------------------------------------------------------------------------------------
%	CHAPTER - The PRINT Statement
%----------------------------------------------------------------------------------------

\chapter*{The PRINT Statement}
\addcontentsline{toc}{chapter}{\protect\numberline{}The PRINT Statement}

Now that we've used the {\ttfamily PRINT} statement to display a string of
letters to the screen, let's look at how it can display numbers.  You can
re-use your existing program by overwriting the statement at line 10.  Do do
this, simply type {\ttfamily 10} followed by the BASIC statement you want to
replace line 10 with.  In this case, let's replace it with a {\ttfamily PRINT}
statement that prints numbers instead of letters:\\

\codeblock {

	10 PRINT "42"\\

}

Make sure you press \returnkey so that the Commander X16 accepts the new line,
and replaces the old one in your program.  Now if you type {\ttfamily RUN} and
hit \returnkey, you should see:

\codeblock{
	42\\
}

This example isn't much different than the last example, because we simply
printed a different string.  The number {\ttfamily 42} is given to the
{\ttfamily PRINT} statement as a string, just as {\ttfamily HELLO, WORLD!} was
given as a string.  To the {\ttfamily PRINT} statement, it doesn't matter which
characters are put inside the quotations marks; it always views them as just a
series of characters.  However, the {\ttfamily PRINT} statement is able to
display a number to the screen while treating it as a number.  For example,
replace line 10 with:\\

\codeblock {

	10 PRINT 42\\

}

Notice that there are no quotation marks.  When type {\ttfamily RUN}, you will
notice a slight difference:\\

\codeblock{
	\hspace*{0.6em}42\\
}

Do you see it?  There is an empty space before the number.  This is because the
{\ttfamily PRINT} statement leaves a space for a sign indicator (positive or
negative) when it prints a number.  For positive numbers nothing is displayed,
but if given a negative number, {\ttfamily PRINT} will fill that space with a
{\ttfamily -}.  Try replacing {\ttfamily 42} with {\ttfamily -42} and running
the program again.\\

Now that you have the {\ttfamily PRINT} statement treating the value as a
number, you can try passing it a mathematical expression that the X16 will {\em
compute} into a number:\\

\codeblock {

	10 PRINT 3+2\\

}

You should see:\\

\codeblock {
	\hspace{0.6em}5\\
}

Now you can compare this result with what happens if we pass {\ttfamily 3+2} to
{\ttfamily PRINT} as a string:\\

\codeblock {

	10 PRINT "3+2"\\

}

This time you should see:\\

\codeblock {
	3+2\\
}

When you pass a value to {\ttfamily PRINT} inside of quotation marks, it treats
it as a series of characters, Specifically "3", "+", and "2".  {\ttfamily
PRINT} is also smart enough to accept a mixture of strings and numeric
values:\\

\codeblock {

	10 PRINT "THERE ARE" 5+3 "BITS IN A BYTE"\\

}

%----------------------------------------------------------------------------------------
%	CHAPTER - Graphic Characters
%----------------------------------------------------------------------------------------

\chapter*{Graphic Characters}
\addcontentsline{toc}{chapter}{\protect\numberline{}Graphic Characters}

Sit down in front of the Commander X16 keyboard and look at the keys.  If
you're familiar with a regular computer or typewriter keyboard, you'll notice
that the X16 keyboard's keys are different.  Rather than just having letters,
numbers, and punctuation on them, you'll see that most keys also have one or
more "graphic characters" printed on them.  While the regular characters can be
typed by simply pressing the keys, the graphics characters can be typed by
holding down either \shiftkey or \altkey and pressing one of the keys with
graphics characters.  For example, try holding down the \shiftkey key and
pressing the \keybackgroundcolor{white}\key{s} key.  Instead of an 'S', a heart
should appear on the screen.  Now press and hold down the \altkey key and press
\key{s}.  This time a right-angled graphic character should appear.  Holding
the \shiftkey key causes the graphic character on the right to be typed, and
holding the \altkey key causes the graphic character on the left to be typed.\\

Graphic characters can be used to create larger shapes and images on the
screen.  For example, the graphic characters that can be typed by holding
\altkey and typing \key{a}, \key{s}, \key{z}, and \key{x} can be used to
create a square that is larger than a single character.  If you include the
graphic characters that can be typed by holding \shiftkey and typing \key{c}
and \key{b}, you can draw a rectangle as large as the entire screen!  If you
want your rectangle to have rounded corners, you can use the graphic characters
on the \key{u}, \key{i}, \key{j}, and \key{k}.\\

\tryit{
	Spend some time experimenting with the other graphic characters to see what
	kinds of shapes and designs you can draw.  You can use the arrow keys to
	position the cursor wherever you need to type.\\
}

% To start, sit down in front of the Commander X16 Keyboard and type the
% following:\\
% 
% Hold down the \shiftkey key and press the \clrhomekey key.  Then type
% \keybackgroundcolor{white}\key{N}\key{E}\key{W} and press the \returnkey key.\\
% 
% This will clear any program currently programmed so that the X16 is ready to
% accept a new one.

%----------------------------------------------------------------------------------------
%	CHAPTER - Colors
%----------------------------------------------------------------------------------------

\chapter*{Colors}
\addcontentsline{toc}{chapter}{\protect\numberline{}Colors}

The Commander X16 is also capable of displaying characters in different colors.
If you look at the number keys on the Commander X16 keyboard, you will see that
numbers 1-8 have colors listed next to the numbers.  By holding down either
\altkey or \ctrlkey and pressing one of the numbers with colors listed on them,
you can change the color of the characters you type.  Holding \ctrlkey while
you type the number will select the top color, and holding \altkey while you
type the number will select the bottom color.  This allows for a total of 16
different colors that can be used to type characters:\\

\begin{tabular}{|c|c|c|}
	\hline
	Key & CTRL & ALT\\ \hline
	1 & Black & Orange \\ \hline
	2 & White & Brown \\ \hline
	3 & Red & Light Red \\ \hline
	4 & Cyan & Dark Gray\\ \hline
	5 & Purple & Medium Gray\\ \hline
	6 & Green & Light Green\\ \hline
	7 & Blue & Light Blue\\ \hline
	8 & Yellow & Light Gray\\ \hline
\end{tabular}

\vspace{16pt}

You may also noticed that \key{9} has "RVS ON" and \key{0} has "RVS OFF" where
the colors are listed on the other number keys.  These are "reverse on" and
"reverse off", and can be typed by holding down \ctrlkey while pressing the
number.  When "reverse on" is typed, it switches which color is used for the
character and which one is used for the background.  For example, if you have
the character colors set to white and the background color set to blue, hitting
\ctrlkey + \key{9} will cause the characters you type next to be displayed in
blue with a white background.  Because each character can be printed with the
colors reversed, it's like there are twice as many characters!  This allows for
a lot of creativity in making drawings and designs with a combination of
graphic characters and colors.  To turn off "reverse", just press \ctrlkey +
\key{0}.\\

\section{Setting Colors While PRINTing}

When you are typing the text of a print statement, you can use \ctrlkey and
\altkey with the number keys to change and reverse the colors.  However, the
colors will not display on the screen in their new colors immediately.
Instead, a graphic character will be written to the text of your {\ttfamily
PRINT} statement.  When the {\ttfamily PRINT} statement executes, this graphic
character will tell the Commander X16 to switch to that color as it is printing
the text.  This also works with the "RVS ON" and "RVS OFF" functions.  By using
this technique, you can change the color of the text multiple times (or even
reverse it) all from a single {\ttfamily PRINT} statement.  If you don't want
your text to stay that way, remember to set it back to how you want it before
your program ends!\\

\section{The COLOR Statement}

The colors can also be set by running the {\ttfamily COLOR} statement.  On a
new line, try typing {\ttfamily COLOR 2} and then pressing \returnkey.  You
should see the "READY" prompt printed to the screen in red.  Now anything you
type will also be in red.  This is the same effect as holding \ctrlkey and
pressing \key{3}.  The {\ttfamily COLOR} statement allows you to set the
foreground color to the first 16 colors of the Commander X16's
palette\footnote{A "palette" refers to a set of predefined colors that can be
used on the Commander X16.  More information on palettes will be discussed in
the chapter on Graphics}.  Because of the way computers work, you specify which
color by typing a number from 0-15 instead of 1-16.

The numbers that specify colors with the {\ttfamily COLOR} statement do not
correspond to the number keys pressed to change the color.  This is because
there are only 8 number keys that contain colors, but there are 16 colors to
choose from.  The order is the same, however, if you read all the top color
values of the number keys from left to right and then read all the bottom
values from left to right.  If you start numbering with 0, you get the values
that you can use with the {\ttfamily COLOR} statement:\\

\begin{tabular}{|c|c|c|c|}
	\hline
	Number & Color & Number & Color\\ \hline
	0 & Black & 8 & Orange \\ \hline
	1 & White & 9 & Brown \\ \hline
	2 & Red & 10 & Light Red \\ \hline
	3 & Cyan & 11 & Dark Gray\\ \hline
	4 & Purple & 12 & Medium Gray\\ \hline
	5 & Green & 13 & Light Green\\ \hline
	6 & Blue & 14 & Light Blue\\ \hline
	7 & Yellow & 15 & Light Gray\\ \hline
\end{tabular}

\vspace{16pt}

The {\ttfamily COLOR} statement can also be used to set both the foreground
color and the background color.  The first argument is the color to use for the
foreground and the second argument is the color to use for the background.
Separate the two arguments with a comma, like so:\\

\codeblock {
	COLOR 13,0\\
}

The above call to the {\ttfamily COLOR} statement will set the foreground color
to light green and the background color to black.  Just like the {\ttfamily
PRINT} statement, the {\ttfamily COLOR} statement can be used in a BASIC
program.  When the {\ttfamily COLOR} statement is run in a program, it does not
advance to the next line nor cause the "READY" prompt to be printed.  This
means that the {\ttfamily COLOR} statement can be used to change the foreground
and background colors in between calls to the {\ttfamily PRINT} statement
without printing extra lines that you don't want.  It is also possible to call
the {\ttfamily PRINT} statement in a way where it does not advance the cursor
to the next line, but instead leaves it at the end of the characters that have
just been printed.  In order to use {\ttfamily PRINT} without advancing to the
next line, simply type a ';' at the end of the statement:\\

\codeblock{
	10 PRINT "X";\\
}

Now the next {\ttfamily PRINT} statement in the program will start printing
directly after the "X" that was written to the screen.  If we use the
{\ttfamily COLOR} statement before the next {\ttfamily PRINT} statement, we can
change both the background and foreground colors of characters that are next to
each other on a line:\\

\codeblock{
	10 PRINT "X";\\
	20 COLOR 2,1\\
	30 PRINT "X";\\
}

If the above code is run, it will print two X's next to each other on the
screen, but they will have different foreground and background colors.  The
Commander X16 allows for each character on the screen to have its own
foreground and background color.  A short program can be written to demonstrate
this:\\

\keytextcolor{black}
\keybackgroundcolor{white}

\codeblock{
	10 COLOR RND(1)*16,RND(1)*16\\
	20 PRINT "X\doublekey{RVS\\ON}X\doublekey{RVS\\OFF}";\\
	30 GOTO 10\\
}

\reminder{
	Did you remember to type {\ttfamily NEW} before entering in a new program?
}

When run, this program will fill up the screen with pairs of X's.  It will
first print each X normally, followed by another X using the same foreground
and background colors, but reversed.\\

%----------------------------------------------------------------------------------------
%CHAPTER - Screen Modes
%----------------------------------------------------------------------------------------

\chapter*{Screen Modes}\index{Screen Modes}\index{SCREEN}
\addcontentsline{toc}{chapter}{\protect\numberline{}Screen Modes}

When you turn on your Commander X16, the screen displays 80 columns across and
60 rows down.  That's a lot of characters!  But what if you want bigger text
that's easier to read from across the room?  The Commander X16 can display text
in several different screen modes, and it's easy to switch between them.\\

\section{Switching Between 80 and 40 Column Modes}

The two most commonly used screen modes are:\\

\begin{itemize}
	\item {\bfseries 80 columns} --- Lots of text fits on the screen (80
		characters wide, 60 rows tall)
	\item {\bfseries 40 columns} --- Larger text that's easier to see (40
		characters wide, 30 rows tall)
\end{itemize}

The easiest way to switch between these two modes is to press the \key{F4} key
and then press \returnkey.  Try it now!  Your screen will clear and the text
will appear much larger.  Press \key{F4} and \returnkey again to go back to 80
column mode.\\

\note{
	You must press \returnkey after pressing \key{F4}.  This prevents you from
	accidentally switching modes when you didn't mean to!\\
}

\section{The SCREEN Statement}

You can also change screen modes from within a BASIC program using the
{\ttfamily SCREEN} statement.  Try this example:\\

\codeblock{
	10 SCREEN 3\\
	20 PRINT "THIS IS 40 COLUMN MODE!"\\
	30 PRINT "THE TEXT IS NICE AND BIG."\\
	40 SLEEP 120\\
	50 SCREEN 0\\
	60 PRINT "NOW WE'RE BACK TO 80 COLUMNS."\\
}

When you {\ttfamily RUN} this program, it switches to 40 column mode, prints a
message, waits two seconds, and then switches back to 80 column mode.  The
{\ttfamily SCREEN} statement clears the screen when it changes modes, so
anything that was displayed before will be erased.\\

\section{Available Screen Modes}

The Commander X16 supports many different text modes.  Here are the most useful
ones:\\

\begin{tabular}{|c|c|l|}
	\hline
	{\bfseries Mode} & {\bfseries Size} & {\bfseries Description}\\ \hline
	0 & 80 $\times$ 60 & Default mode, lots of room for text\\ \hline
	1 & 80 $\times$ 30 & Wide but shorter\\ \hline
	2 & 40 $\times$ 60 & Big text, tall screen\\ \hline
	3 & 40 $\times$ 30 & Big text, like classic computers\\ \hline
\end{tabular}

\vspace{16pt}

Mode 3 (40 $\times$ 30) is especially nice if you're connecting your X16 to an
older television or CRT monitor, because the larger characters are easier to
read on those displays.\\

\tryit{
	Write a program that prints your name in each of the four screen modes
	listed above.  Use {\ttfamily SLEEP 60} between each mode change so you
	have time to see the result!\\
}

\section{A Quick Mode Toggle}

There's a handy shortcut for switching between 80 and 40 column modes.  Using
{\ttfamily SCREEN -1} will toggle between the two modes:\\

\codeblock{
	10 SCREEN -1\\
	20 PRINT "PRESS A KEY TO TOGGLE"\\
	30 GET A\$: IF A\$="" THEN 30\\
	40 GOTO 10\\
}

This program toggles between 40 and 80 column mode each time you press a key.
It's a fun way to see the difference between the two modes!\\

\section{Starting Up in Your Favorite Mode}

If you prefer to always start your Commander X16 in 40 column mode, you can do
this by modifying your AUTOBOOT.X16 file.  This is an advanced topic, but
here's a hint: by adding {\ttfamily SCREEN 3} at the beginning of your
AUTOBOOT.X16 file, your X16 will switch to 40 column mode every time it starts
up.\\

But wait --- how do you modify a file that already exists?  That's exactly what
we'll learn in the next section!

%----------------------------------------------------------------------------------------
%	CHAPTER - Editing Text
%----------------------------------------------------------------------------------------

\chapter*{Editing Text}\index{Editing}\index{Cursor}\index{Full Screen Editor}
\addcontentsline{toc}{chapter}{\protect\numberline{}Editing Text}

When you turn on your Commander X16, it starts up in the \emph{full screen
editor}.  This is different from many other computers where you can only type
at the bottom of the screen.  On the X16, you can move the cursor anywhere on
the screen --- up, down, left, or right --- and type or edit text wherever you
want!\\

This is incredibly useful because it means you can go back and fix mistakes, or
even re-run commands you typed earlier.  If you typed something wrong and got
an error, you don't have to type the whole thing again.  Just move the cursor
back up to the line with the mistake, fix it, and press \returnkey.  Let's
learn how this works on this flubbed classic program:\\


\begin{center}
	\screenbox{2.75in}{2in}{

		NEW\\
		
		READY.\\
		10 PIRNT7 "HE WORLD"\\
		RUN\\

		?SYNTAX ERROR IN 10\\
		READY.\\
		\cursor

	}
\end{center}

\section{The Cursor Keys}

You've already seen the blinking square called the \emph{cursor}.  The cursor
shows you where the next character you type will appear on the screen.  You can
move the cursor anywhere on the screen using the arrow keys.\\

Try pressing the arrow keys now.  You'll find four of them on the right side of
the keyboard:\\

\begin{itemize}
	\item \key{$\uparrow$} moves the cursor up one line
	\item \key{$\downarrow$} moves the cursor down one line
	\item \key{$\leftarrow$} moves the cursor left one space
	\item \key{$\rightarrow$} moves the cursor right one space
\end{itemize}

If you hold down an arrow key, the cursor keeps moving in that direction until
you let go.  This makes it easy to move quickly across the screen!\\

\section{Fixing Mistakes Without Retyping}

Let's see the power of the full screen editor in action.  First type in this line
exactly as shown (with the mistakes!)\\

\codeblock{
	NEW\\
	10 PIRNT7 "HE WORLD"\\
	RUN\\
}

You should see an error message:\\

\codeblock{
	?SYNTAX ERROR IN 10\\
	READY.\\
}

The syntax error is because we didn't type in a valid BASIC statement: we typed
{\ttfamily PIRNT7} instead of {\ttfamily PRINT}.  Even inside the quotes there
is a second typo --- we wrote {\ttfamily HE WORLD} instead of {\ttfamily HELLO
WORLD}.  On most computers, you would have to type the whole line again.  But
not on the Commander X16!\\

\subsection{Using the Cursor Keys to Go Back}

Use the \key{$\uparrow$} arrow key to move the cursor back up to the line 10 you
typed.  Position the cursor so it's on top of the {\ttfamily I} in {\ttfamily
PIRNT7}.  Now type {\ttfamily R} --- it replaces the {\ttfamily I}.  Then type
{\ttfamily I} --- it replaces the {\ttfamily N}.  Now you have {\ttfamily
PRINT7} which is closer, but we still need to get rid of the extra {\ttfamily 7}!\\

\subsection{Using Backspace to Remove Extra Characters}

Use the cursor keys to move to the right, stopping on the space after the
{\ttfamily 7}.  Now press the \backspacekey key.  The {\ttfamily 7} disappears
and the rest of the line shifts left.  Now you have {\ttfamily 10 PRINT "HE
WORLD"}.  Almost there!\\

\subsection{Using Insert to Add Missing Characters}

Use the cursor keys to move to the right, stopping on the space between
{\ttfamily HE} and {\ttfamily WORLD}.  We can't just type here, because it
would overwrite the rest of our string --- we don't have enough space!  Never
fear!  We can use the \insertkey key!\\

Press the \insertkey key.  Notice that the entire rest of the line moves to the
right.  The cursor is now waiting for the character we want to put there.  Type
\key{L}.  Hmmm, but we still need more room.  Press the \insertkey key two more
times and type \key{L}\key{O}.  The line should now read the familiar
{\ttfamily 10 PRINT "HELLO WORLD"}.  We've corrected all the mistakes without
typing the entire line again!\\

With the cursor anywhere on the line, press \returnkey to accept the changes.
Now your cursor has moved down to the line with your previous {\ttfamily RUN}
command.  Press \returnkey again to run your fixed program.  You should see:\\

\begin{center}
	\screenbox{2.75in}{2in}{

		NEW\\
		
		READY.\\
		10 PRINT "HELLO WORLD"\\
		RUN\\
		HELLO WORLD\\
		?SYNTAX ERROR IN 10\\
		READY.\\
		\cursor

	}
\end{center}

Wait --- why does it still say {\ttfamily ?SYNTAX ERROR IN 10}?  Don't worry!
Look carefully: your correct {\ttfamily HELLO WORLD} message is printed
\emph{above} the old error.  The error message is still visible because we
never cleared the screen --- it's just leftover text from our earlier failed
attempt.  Your program is working perfectly!\\

This can be confusing at first, so remember: the Commander X16's screen is like
a chalkboard.  Old text stays visible until you erase it.  To clean up the
display, hold down \shiftkey and press \clrhomekey to clear the screen.\\

\note{
	The \backspacekey key deletes the character to the \emph{left} of the
	cursor (and moves the cursor back), while the \deletekey key deletes the
	character \emph{under} the cursor (and pulls everything to the right
	closer).  Both are useful for different situations!\\
}

\section{The Home Key}

The \clrhomekey key has two useful functions:\\

\begin{itemize}
	\item Pressing \clrhomekey by itself moves the cursor to the ``home''
		position --- the upper-left corner of the screen.
	\item Pressing \shiftkey + \clrhomekey clears the entire screen and moves
		the cursor to the home position.
\end{itemize}

This is a quick way to get a fresh start without losing your program.
Remember, clearing the screen doesn't erase your program from memory --- it
just cleans up the display.\\

\section{Saving Your Work to a File}

When you type a program into the Commander X16, it lives in the computer's
memory.  But if you turn off the computer, that memory is erased and your
program is gone!  To keep your program, you need to \emph{save} it to the SD
card as a file.\\

Let's save a simple program.  First, clear any old program by typing
{\ttfamily NEW} and pressing \returnkey.  Now type this one-line program:\\

\codeblock{
	10 PRINT "HELLO"\\
}

To save this program to a file, use the {\ttfamily SAVE} command followed by a
filename in quotation marks:\\

\codeblock{
	SAVE "HELLO.BAS"\\
}

You should see {\ttfamily SAVING HELLO.BAS} followed by {\ttfamily READY.}
Your program is now safely stored on the SD card!\\

\section{Loading a Program from a File}

Later, when you want to work on your program again, you can \emph{load} it back
into memory using the {\ttfamily LOAD} command:\\

\codeblock{
	LOAD "HELLO.BAS"\\
}

After loading, type {\ttfamily LIST} to see your program, or {\ttfamily RUN} to
run it.\\

\section{Deleting a File}

If you no longer need a file, you can delete it using the {\ttfamily SCRATCH}
command (through the DOS command interface):\\

\codeblock{
	DOS "S:HELLO.BAS"\\
}

Be careful with this command --- once a file is deleted, it's gone forever!\\

\section{Editing a Saved Program}

Now let's put your editing skills to use on a real file.  Remember how we
mentioned modifying the AUTOBOOT.X16 file to change your startup screen mode?
Here's how you would do it.\\

First, let's see if you have an AUTOBOOT.X16 file.  Type:\\

\codeblock{
	LOAD "AUTOBOOT.X16"\\
}

If you see {\ttfamily FILE NOT FOUND ERROR}, that's okay --- it just means you
don't have one yet.  In that case, we'll create a simple one.  Type:\\

\codeblock{
	NEW\\
	10 REM MY AUTOBOOT FILE\\
}

If you did have an AUTOBOOT.X16 file and it loaded successfully, type
{\ttfamily LIST} to see what's in it.  Now let's add a line to switch to 40
column mode at startup.  Type:\\

\codeblock{
	5 SCREEN 3\\
}

Because we used line number 5, this new line will appear \emph{before} any
existing lines in the program.  Type {\ttfamily LIST} to see your program with
the new line added.\\

To save your changes, type:\\

\codeblock{
	SAVE "AUTOBOOT.X16"\\
}

If you already had an AUTOBOOT.X16 file, use this command instead to overwrite
the old version:\\

\codeblock{
	SAVE "@:AUTOBOOT.X16"\\
}

The next time you turn on your Commander X16, it will automatically run this
program and switch to 40 column mode!\\

\tryit{
	Practice your editing skills by typing in a short program with some
	intentional mistakes, then use the cursor keys, \insertkey, and \deletekey
	to fix them without retyping the whole line.\\
}

\@openrighttrue\makeatother
